---
title: "Evaluating Alert Fatigue Codebook"
author: "Brianna Jackson"
date: "2026-03-30"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/brian/OneDrive/Documents")
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 8,
  fig.height = 5,
  dpi = 150
)

library(data.table)
library(MASS)
library(ggplot2)
library(lubridate)
library(tidyr)
library(scales)
library(knitr)
library(kableExtra)
library(broom) 
library(dplyr)  # always last
```

```{r load-data, echo=FALSE}
cohort <- fread("C:/Users/brian/OneDrive/Documents/mimic_icustay_cohort_final.csv")

prescriptions <- fread(
  "C:/Users/brian/OneDrive/Documents/prescriptions.csv.gz",
  select = c("subject_id", "hadm_id", "starttime", "stoptime", "drug")
)
prescriptions$starttime <- as.POSIXct(prescriptions$starttime)
cohort$intime           <- as.POSIXct(cohort$intime)

cohort_lookup <- cohort %>%
  dplyr::select(hadm_id, intime, icu_los_days) %>%
  distinct(hadm_id, .keep_all = TRUE)

prescriptions_cohort <- prescriptions %>%
  dplyr::filter(hadm_id %in% cohort$hadm_id, !is.na(starttime)) %>%
  left_join(cohort_lookup, by = "hadm_id") %>%
  mutate(
    order_hour        = as.integer(format(starttime, "%H")),
    hours_since_admit = as.numeric(difftime(starttime, intime, units = "hours")),
    day_of_stay       = floor(hours_since_admit / 24) + 1,
    shift = factor(case_when(
      order_hour >= 7  & order_hour < 15 ~ "Day (7am-3pm)",
      order_hour >= 15 & order_hour < 23 ~ "Evening (3pm-11pm)",
      TRUE                               ~ "Night (11pm-7am)"
    ), levels = c("Day (7am-3pm)", "Evening (3pm-11pm)", "Night (11pm-7am)"))
  ) %>%
  dplyr::filter(hours_since_admit >= 0)

hourly_counts <- prescriptions_cohort %>%
  group_by(order_hour) %>%
  summarise(n_orders = n(), .groups = "drop") %>%
  mutate(pct = n_orders / sum(n_orders) * 100)

top_hours <- hourly_counts %>% arrange(desc(n_orders)) %>% slice(1:3)

shift_summary <- prescriptions_cohort %>%
  group_by(shift) %>%
  summarise(
    n_orders = n(),
    pct      = round(n() / nrow(prescriptions_cohort) * 100, 1),
    .groups  = "drop"
  )

daily_burden <- prescriptions_cohort %>%
  dplyr::filter(day_of_stay >= 1, day_of_stay <= 14) %>%
  group_by(day_of_stay) %>%
  summarise(
    n_orders       = n(),
    n_admissions   = n_distinct(hadm_id),
    orders_per_adm = n_orders / n_admissions,
    .groups = "drop"
  )

burst_metric <- cohort %>%
  dplyr::select(hadm_id, max_burst_60min, burst_category) %>%
  dplyr::filter(!is.na(max_burst_60min))

# Winsorize lab events at 99th percentile (as documented in preprocessing notes)
p99_lab <- quantile(cohort$n_lab_events, 0.99, na.rm = TRUE)
cohort <- cohort %>%
  mutate(n_lab_events_w = pmin(n_lab_events, p99_lab))

scenario_long <- data.frame(
  Events   = c(cohort$total_events, cohort$scenario1_events,
               cohort$scenario2_events, cohort$scenario3_events),
  Scenario = rep(c("Baseline", "S1: Rx Dedup -15%",
                   "S2: Lab Dedup -20%", "S3: Combined"), each = nrow(cohort))
) %>%
  dplyr::filter(Events < 3000) %>%
  mutate(Scenario = factor(Scenario,
    levels = c("Baseline", "S1: Rx Dedup -15%",
               "S2: Lab Dedup -20%", "S3: Combined")))
```

------------------------------------------------------------------------

# Dataset Overview

```{r overview-table}
overview <- data.frame(
  Item  = c("Data Source", "Unit of Analysis", "Total Observations",
            "Total Variables", "Date Constructed", "Software"),
  Value = c(
    "MIMIC-IV (PhysioNet / Beth Israel Deaconess Medical Center)",
    "ICU stay (stay_id), linked to hospital admission (hadm_id)",
    format(nrow(cohort), big.mark = ","),
    ncol(cohort),
    "March 2026",
    "R 4.x with data.table, dplyr, MASS, DuckDB, lubridate"
  )
)

kable(overview, col.names = c("", ""), booktabs = TRUE,
      caption = "Table 1. Dataset Summary") %>%
  kable_styling(full_width = FALSE, position = "left")
```

------------------------------------------------------------------------

# Variable Codebook

```{r codebook-table}
codebook <- data.frame(
  Variable = c(
    "subject_id", "hadm_id", "stay_id",
    "anchor_age", "gender", "first_careunit",
    "admittime", "dischtime", "intime", "outtime",
    "hospital_los_days", "icu_los_hours", "icu_los_days",
    "n_prescriptions", "n_unique_drugs", "rx_per_day",
    "n_lab_events", "n_unique_labitems", "n_abnormal_labs", "labs_per_day",
    "total_events", "total_events_per_day",
    "scenario1_events", "scenario2_events", "scenario3_events",
    "max_burst_60min", "burst_category"
  ),
  Type = c(
    "Integer", "Integer", "Integer",
    "Integer", "Factor", "Factor",
    "POSIXct", "POSIXct", "POSIXct", "POSIXct",
    "Numeric", "Numeric", "Numeric",
    "Integer", "Integer", "Numeric",
    "Integer", "Integer", "Integer", "Numeric",
    "Integer", "Numeric",
    "Integer", "Integer", "Integer",
    "Integer", "Factor"
  ),
  Source = c(
    rep("MIMIC-IV: icustays / patients / admissions", 3),
    rep("MIMIC-IV: patients", 2), "MIMIC-IV: icustays",
    rep("MIMIC-IV: admissions", 2), rep("MIMIC-IV: icustays", 2),
    rep("Derived", 3),
    rep("MIMIC-IV: prescriptions (aggregated)", 2), "Derived",
    rep("MIMIC-IV: labevents (aggregated)", 3), "Derived",
    rep("Derived", 2),
    rep("Simulated (scenario analysis)", 3),
    rep("Derived: prescriptions timestamps", 2)
  ),
  Description = c(
    "Unique patient identifier",
    "Hospital admission ID; primary join key across tables",
    "ICU stay identifier",
    "Patient age at MIMIC-IV anchor year",
    "Patient sex (M = Male, F = Female)",
    "ICU unit type at time of admission",
    "Hospital admission timestamp",
    "Hospital discharge timestamp",
    "ICU admission timestamp",
    "ICU discharge timestamp",
    "Hospital length of stay in days (dischtime - admittime)",
    "ICU length of stay in hours (outtime - intime)",
    "ICU length of stay in days (icu_los_hours / 24)",
    "Total medication prescription orders per hospital admission",
    "Count of distinct drug names ordered per hospital admission",
    "Medication orders per ICU day (n_prescriptions / icu_los_days)",
    "Total laboratory measurements per hospital admission",
    "Count of distinct laboratory item types per admission",
    "Laboratory measurements flagged as abnormal",
    "Lab events per ICU day (n_lab_events / icu_los_days)",
    "Total CDS-triggering events (n_prescriptions + n_lab_events)",
    "Total events per ICU day (total_events / icu_los_days)",
    "Simulated burden: Rx orders reduced 15% (deduplication scenario)",
    "Simulated burden: Lab events reduced 20% (suppression scenario)",
    "Simulated burden: Combined Rx -15% and Lab -20%",
    "Maximum medication orders in any 60-minute window per admission",
    "Burst intensity category: Low / Moderate / High / Very High"
  )
)

kable(codebook, booktabs = TRUE, longtable = TRUE,
      caption = "Table 2. Variable Codebook — All 27 Variables") %>%
  kable_styling(font_size = 10) %>%
  column_spec(1, bold = TRUE, width = "1.6in") %>%
  column_spec(2, width = "0.7in") %>%
  column_spec(3, width = "1.6in") %>%
  column_spec(4, width = "3in")
```

------------------------------------------------------------------------

# Descriptive Statistics

```{r desc-stats}
desc <- cohort %>%
  dplyr::summarise(
    across(
      c(n_prescriptions, n_unique_drugs, rx_per_day,
        n_lab_events, n_unique_labitems, n_abnormal_labs,
        total_events, total_events_per_day,
        max_burst_60min, icu_los_days, anchor_age),
      list(
        N      = ~sum(!is.na(.)),
        Mean   = ~round(mean(., na.rm = TRUE), 1),
        SD     = ~round(sd(., na.rm = TRUE), 1),
        Median = ~round(median(., na.rm = TRUE), 1),
        Q1     = ~round(quantile(., 0.25, na.rm = TRUE), 1),
        Q3     = ~round(quantile(., 0.75, na.rm = TRUE), 1),
        Min    = ~round(min(., na.rm = TRUE), 1),
        Max    = ~round(max(., na.rm = TRUE), 1)
      )
    )
  ) %>%
  tidyr::pivot_longer(everything(),
    names_to  = c("Variable", "Stat"),
    names_sep = "_(?=[^_]+$)",
    values_to = "Value"
  ) %>%
  tidyr::pivot_wider(names_from = Stat, values_from = Value)

kable(desc, booktabs = TRUE,
      caption = "Table 3. Descriptive Statistics: Burden and Temporal Variables") %>%
  kable_styling(font_size = 9)
```

```{r cohort-demographics}
#cohort-demographics
#Gender split

cohort %>%
  group_by(gender) %>%
  summarise(
    n   = n(),
    pct = round(n() / nrow(cohort) * 100, 1)
  ) %>%
  kable(col.names = c("Gender", "N", "%"),
        caption   = "Table 3a. Gender Distribution",
        booktabs  = TRUE) %>%
  kable_styling(full_width = FALSE)

# LOS summary
cohort %>%
  summarise(
    median_los = round(median(icu_los_days, na.rm = TRUE), 1),
    mean_los   = round(mean(icu_los_days,   na.rm = TRUE), 1),
    q1_los     = round(quantile(icu_los_days, 0.25, na.rm = TRUE), 1),
    q3_los     = round(quantile(icu_los_days, 0.75, na.rm = TRUE), 1),
    max_los    = round(max(icu_los_days,    na.rm = TRUE), 1)
  ) %>%
  kable(col.names = c("Median LOS", "Mean LOS", "Q1", "Q3", "Max"),
        caption   = "Table 3b. ICU Length of Stay Summary (days)",
        booktabs  = TRUE) %>%
  kable_styling(full_width = FALSE)
```

------------------------------------------------------------------------

# Statistical Approach

## Why Negative Binomial Regression?

The outcome variables in this analysis are counts — whole numbers that
can never be negative (medication orders, lab tests, total events).
Standard linear regression is not appropriate because it can predict
impossible negative values and assumes a normal distribution that count
data does not follow.

The simpler count model — Poisson regression — also does not fit here
because it requires the variance to equal the mean. Our data is far more
spread out than Poisson can handle. This is called **overdispersion**.

```{r overdispersion-check}
od <- data.frame(
  Variable = c("Medication orders", "Lab events", "Total events"),
  Mean     = round(c(mean(cohort$n_prescriptions, na.rm = TRUE),
                     mean(cohort$n_lab_events,    na.rm = TRUE),
                     mean(cohort$total_events,    na.rm = TRUE)), 1),
  Variance = round(c(var(cohort$n_prescriptions,  na.rm = TRUE),
                     var(cohort$n_lab_events,     na.rm = TRUE),
                     var(cohort$total_events,     na.rm = TRUE)), 1),
  Ratio    = round(c(var(cohort$n_prescriptions,  na.rm = TRUE) /
                       mean(cohort$n_prescriptions, na.rm = TRUE),
                     var(cohort$n_lab_events,     na.rm = TRUE) /
                       mean(cohort$n_lab_events,    na.rm = TRUE),
                     var(cohort$total_events,     na.rm = TRUE) /
                       mean(cohort$total_events,    na.rm = TRUE)), 0)
)

kable(od,
      col.names = c("Variable", "Mean", "Variance", "Variance / Mean"),
      caption   = "Table 4. Overdispersion Check.",
      booktabs  = TRUE) %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(4, bold = TRUE)
```

Negative binomial regression handles overdispersion by adding a
dispersion parameter (theta) that absorbs the extra spread in the data.
This produces honest standard errors and reliable confidence intervals.

## Formal Confirmation: Poisson vs. Negative Binomial

```{r lrt-test}
cohort_test <- cohort %>%
  dplyr::filter(icu_los_days > 0, !is.na(anchor_age), !is.na(first_careunit), !is.na(gender),
    !is.na(n_prescriptions),
    !is.na(n_lab_events_w),
    !is.na(total_events))

nb_test <- glm.nb(
  n_prescriptions ~ anchor_age + gender + first_careunit +
    offset(log(icu_los_days)), data = cohort_test
)
pois_test <- glm(
  n_prescriptions ~ anchor_age + gender + first_careunit +
    offset(log(icu_los_days)), family = poisson, data = cohort_test
)

lrt   <- 2 * (logLik(nb_test) - logLik(pois_test))
lrt_p <- pchisq(as.numeric(lrt), df = 1, lower.tail = FALSE)

data.frame(
  Comparison = "Poisson vs. Negative Binomial",
  ChiSq      = format(round(as.numeric(lrt), 0), big.mark = ","),
  p_value    = format(lrt_p, scientific = TRUE),
  NB_Theta   = round(nb_test$theta, 3),
  Decision   = "Negative Binomial"
) %>%
  kable(col.names = c("Comparison", "Chi-Squared", "p-value",
                      "NB Theta", "Decision"),
        caption   = "Table 5. Likelihood Ratio Test.",
        booktabs  = TRUE) %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(5, bold = TRUE)
```

## What the Offset Term Does

All models include `offset(log(icu_los_days))`. Without this, a 30-day
stay would appear to have higher burden than a 2-day stay simply because
of duration, not intensity. The offset converts the model to predict
**events per ICU day**, making every encounter comparable regardless of
how long the stay was.

## How to Read the Results (IRR)

```{r irr-guide, echo=FALSE}
data.frame(
  IRR     = c("1.0", "1.20", "0.81", "1.035", "0.189"),
  Meaning = c("No difference from the reference group",
              "20% MORE events per day than the reference group",
              "19% FEWER events per day than the reference group",
              "3.5% more events per day — small but real",
              "81% FEWER events per day — very large difference"),
  Example = c("CVICU — our reference unit in Models 1 and 3",
              "MICU/SICU vs. CVICU",
              "Example only",
              "Male vs. Female patients",
              "Neuro Intermediate vs. CVICU")
) %>%
  kable(col.names = c("IRR", "What It Means", "From Our Study"),
        caption   = "Table 6. How to Read an Incidence Rate Ratio (IRR).",
        booktabs  = TRUE) %>%
  kable_styling(font_size = 10) %>%
  column_spec(1, bold = TRUE)
```

## Key Results: Forest Plot

```{r irr-forest, fig.height=6}
data.frame(
  Unit = c("MICU/SICU","PACU","CVICU (ref)","SICU","Trauma SICU",
           "CCU","MICU","Neuro SICU","Neuro Stepdown","Neuro Intermediate"),
  IRR  = c(1.202,1.659,1.000,0.933,0.852,0.859,0.855,0.522,0.418,0.189),
  Low  = c(1.168,1.344,1.000,0.906,0.826,0.833,0.833,0.492,0.392,0.182),
  High = c(1.236,2.077,1.000,0.960,0.878,0.885,0.877,0.555,0.447,0.196)
) %>%
  mutate(
    Unit      = reorder(Unit, IRR),
    Direction = case_when(IRR > 1  ~ "More than CVICU",
                          IRR == 1 ~ "Reference",
                          TRUE     ~ "Less than CVICU")
  ) %>%
  ggplot(aes(x = IRR, y = Unit, color = Direction)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50", linewidth = 0.8) +
  geom_errorbarh(aes(xmin = Low, xmax = High), height = 0.3, linewidth = 0.8) +
  geom_point(size = 3.5) +
  scale_color_manual(values = c("More than CVICU" = "#C0392B",
                                "Reference"       = "gray40",
                                "Less than CVICU" = "#2E5090")) +
  labs(title   = "Which ICU Units Have Higher or Lower Daily Alert Burden?",
       x       = "Incidence Rate Ratio (IRR)",
       y       = NULL, color = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

```{r model1-rx}

m1 <- glm.nb(
  n_prescriptions ~ anchor_age + gender + first_careunit +
    offset(log(icu_los_days)),
  data = cohort_test
)

m1_tidy <- broom::tidy(m1) %>%
  mutate(
    IRR = exp(estimate),
    conf.low = exp(estimate - 1.96 * std.error),
    conf.high = exp(estimate + 1.96 * std.error),
    across(c(IRR, conf.low, conf.high), ~ round(., 3)),
    std.error = round(std.error, 3),
    statistic = round(statistic, 3),
    p.value = round(p.value, 4)
  ) %>%
  select(term, IRR, std.error, statistic, p.value, conf.low, conf.high)

kable(
  m1_tidy,
  col.names = c("Predictor", "IRR", "Std. Error",
                "Statistic", "p-value",
                "95% CI Low", "95% CI High"),
  caption = "Table 7. Model 1 — Medication Orders per Day",
  booktabs = TRUE
) %>%
  kable_styling(font_size = 9)

```

```{r model2-lab}

m2 <- glm.nb(
  n_lab_events_w ~ anchor_age + gender + first_careunit +
    offset(log(icu_los_days)),
  data = cohort_test
)

m2_tidy <- broom::tidy(m2) %>%
  mutate(
    IRR = exp(estimate),
    conf.low = exp(estimate - 1.96 * std.error),
    conf.high = exp(estimate + 1.96 * std.error),
    across(c(IRR, conf.low, conf.high), ~ round(., 3)),
    std.error = round(std.error, 3),
    statistic = round(statistic, 3),
    p.value = round(p.value, 4)
  ) %>%
  select(term, IRR, std.error, statistic, p.value, conf.low, conf.high)

kable(
  m2_tidy,
  col.names = c("Predictor", "IRR", "Std. Error",
                "Statistic", "p-value",
                "95% CI Low", "95% CI High"),
  caption = "Table 8. Model 2 — Laboratory Events per Day",
  booktabs = TRUE
) %>%
  kable_styling(font_size = 9)
```

```{r model3-total}

m3 <- glm.nb(
  total_events ~ anchor_age + gender + first_careunit +
    offset(log(icu_los_days)),
  data = cohort_test
)

m3_tidy <- broom::tidy(m3) %>%
  mutate(
    IRR = exp(estimate),
    conf.low = exp(estimate - 1.96 * std.error),
    conf.high = exp(estimate + 1.96 * std.error),
    across(c(IRR, conf.low, conf.high), ~ round(., 3)),
    std.error = round(std.error, 3),
    statistic = round(statistic, 3),
    p.value = round(p.value, 4)
  ) %>%
  select(term, IRR, std.error, statistic, p.value, conf.low, conf.high)

kable(
  m3_tidy,
  col.names = c("Predictor", "IRR", "Std. Error",
                "Statistic", "p-value",
                "95% CI Low", "95% CI High"),
  caption = "Table 9. Model 3 — Total CDS Events per Day",
  booktabs = TRUE
) %>%
  kable_styling(font_size = 9)
```

# Visualizations

## Distribution of Total CDS-Triggering Events

```{r plot-total-dist, fig.cap="Figure 2. Distribution of total CDS-triggering events per ICU admission. Medication orders and laboratory measurements combined. X-axis truncated at 3,000; right tail extends to 22,000+. MIMIC-IV, n = 94,444 ICU stays."}
ggplot(cohort, aes(x = total_events)) +
  geom_histogram(bins = 60, fill = "#2E5090", color = "white", alpha = 0.85) +
  scale_x_continuous(labels = comma, limits = c(0, 3000)) +
  labs(
    title    = "Distribution of Total CDS-Triggering Events per ICU Admission",
    subtitle = "Medication orders + laboratory measurements | MIMIC-IV (n = 94,444)",
    x = "Total Events per Admission", y = "Number of ICU Admissions",
    caption = "X-axis truncated at 3,000. Right tail extends to 22,000+."
  ) +
  theme_minimal(base_size = 13)
```

## Event Rate by ICU Unit Type

```{r plot-unit-boxplot, fig.height=6, fig.cap="Figure 3. Daily CDS-triggering event rate by ICU unit type. Events per ICU day (medication orders + lab measurements). Y-axis truncated at 500 for readability. MIMIC-IV."}
ggplot(cohort %>% dplyr::filter(!is.na(first_careunit)),
       aes(x = reorder(first_careunit, total_events_per_day, median, na.rm = TRUE),
           y = total_events_per_day, fill = first_careunit)) +
  geom_boxplot(outlier.alpha = 0.2, outlier.size = 0.6) +
  coord_flip() +
  scale_y_continuous(labels = comma, limits = c(0, 500)) +
  labs(
    title    = "Daily CDS-Triggering Event Rate by ICU Unit Type",
    subtitle = "Events per ICU day | MIMIC-IV",
    x = NULL, y = "Events per ICU Day",
    caption = "Y-axis truncated at 500 for readability."
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

## Medication Orders vs. Lab Events

```{r plot-rx-vs-lab, fig.cap="Figure 4. Relationship between medication orders and laboratory events per ICU admission. Each point represents one ICU stay. Axes truncated for readability. Orange line shows OLS trend with 95% confidence interval. MIMIC-IV."}
ggplot(cohort %>% dplyr::filter(n_prescriptions < 500, n_lab_events < 2000),
       aes(x = n_prescriptions, y = n_lab_events)) +
  geom_point(alpha = 0.08, size = 0.6, color = "#2E5090") +
  geom_smooth(method = "lm", color = "#E05A1E", se = TRUE, linewidth = 1) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Medication Orders vs. Laboratory Events per Admission",
    subtitle = "Each point = one ICU stay | axes truncated for readability",
    x = "Prescription Orders (n)", y = "Laboratory Events (n)",
    caption = "Orange line = OLS trend with 95% CI."
  ) +
  theme_minimal(base_size = 13)
```

## Suppression Scenario Comparison

```{r plot-scenarios, fig.cap="Figure 5. Estimated CDS event burden under three suppression scenarios compared to baseline. S1 = medication order deduplication (-15%); S2 = laboratory event suppression (-20%); S3 = combined strategy. Events truncated at 3,000 for readability. MIMIC-IV, n = 94,444."}
ggplot(scenario_long, aes(x = Events, fill = Scenario)) +
  geom_density(alpha = 0.55) +
  scale_fill_manual(values = c("#1F3864", "#2E75B6", "#E05A1E", "#70AD47")) +
  labs(
    title    = "Estimated CDS Event Burden Under Suppression Scenarios",
    subtitle = "Density of total events per admission | MIMIC-IV",
    x = "Total CDS-Triggering Events per Admission",
    y = "Density", fill = NULL,
    caption = "Events truncated at 3,000 for readability."
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

## ICU Length of Stay Distribution

```{r plot-los, fig.cap="Figure 6. ICU length of stay distribution. Stays exceeding 30 days (less than 1% of cohort) excluded for readability. MIMIC-IV, n = 94,444."}
ggplot(cohort %>% dplyr::filter(icu_los_days <= 30), aes(x = icu_los_days)) +
  geom_histogram(bins = 60, fill = "#2E5090", color = "white", alpha = 0.85) +
  labs(
    title    = "ICU Length of Stay Distribution",
    subtitle = "Stays <= 30 days shown | MIMIC-IV",
    x = "ICU Length of Stay (days)", y = "Number of ICU Admissions",
    caption = "Stays > 30 days (n < 1%) excluded for readability."
  ) +
  theme_minimal(base_size = 13)
```

------------------------------------------------------------------------

# Temporal Clustering Visualizations

## Medication Orders by Hour of Day

```{r plot-hourly, fig.cap="Figure 7. Medication order activity by hour of day across all ICU cohort admissions. Orange bars indicate the three peak hours: 8am (10.7%), 8pm (9.0%), and 10am (7.2%). MIMIC-IV, 8,578,568 total orders."}
ggplot(hourly_counts, aes(x = order_hour, y = n_orders)) +
  geom_col(fill = "#2E5090", alpha = 0.85) +
  geom_col(
    data = hourly_counts %>% dplyr::filter(order_hour %in% top_hours$order_hour),
    fill = "#E05A1E", alpha = 0.9
  ) +
  scale_x_continuous(
    breaks = seq(0, 23, 2),
    labels = c("12am","2am","4am","6am","8am","10am",
               "12pm","2pm","4pm","6pm","8pm","10pm")
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Medication Order Activity by Hour of Day",
    subtitle = "Orange bars = top 3 peak hours | MIMIC-IV ICU cohort",
    x = "Hour of Day", y = "Total Medication Orders",
    caption = "Peak hours: 8am (10.7%), 8pm (9.0%), 10am (7.2%)."
  ) +
  theme_minimal(base_size = 13)
```

## Orders by Clinical Shift

```{r plot-shift, fig.cap="Figure 8. Medication order volume by clinical shift. Day and evening shifts together account for 81.4% of all orders. MIMIC-IV ICU cohort."}
ggplot(shift_summary, aes(x = shift, y = n_orders, fill = shift)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = paste0(pct, "%")),
            vjust = -0.5, size = 4.5, fontface = "bold") +
  scale_fill_manual(values = c("#2E5090", "#E05A1E", "#1F3864")) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "Medication Order Volume by Clinical Shift",
    subtitle = "MIMIC-IV ICU cohort",
    x = NULL, y = "Total Medication Orders",
    caption = "Day + Evening shifts account for 81.4% of all orders."
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

## Orders by Day of ICU Stay

```{r plot-dayofstay, fig.cap="Figure 9. Average medication orders per admission by ICU day of stay. Days 1 through 14 shown. Day 1 corresponds to ICU admission day. Orange line shows loess smoothing trend. MIMIC-IV ICU cohort."}
ggplot(daily_burden, aes(x = day_of_stay, y = orders_per_adm)) +
  geom_line(color = "#2E5090", linewidth = 1.2) +
  geom_point(color = "#2E5090", size = 2.5) +
  geom_smooth(method = "loess", color = "#E05A1E",
              se = TRUE, alpha = 0.15, linewidth = 1) +
  scale_x_continuous(breaks = 1:14) +
  labs(
    title    = "Average Medication Orders per Admission by ICU Day of Stay",
    subtitle = "Days 1-14 shown | MIMIC-IV ICU cohort",
    x = "Day of ICU Stay", y = "Mean Orders per Admission",
    caption = "Day 1 = ICU admission day. Orange = loess trend."
  ) +
  theme_minimal(base_size = 13)
```

## 60-Minute Burst Distribution

```{r plot-burst, fig.height=5, fig.cap="Figure 10. Distribution of maximum medication orders placed within any 60-minute window per ICU admission. Dashed lines mark High (20+) and Very High (30+) thresholds. X-axis truncated at 100; maximum observed value was 432. Median = 18. MIMIC-IV."}
ggplot(burst_metric, aes(x = max_burst_60min)) +
  geom_histogram(bins = 50, fill = "#2E5090", color = "white", alpha = 0.85) +
  geom_vline(xintercept = 20, linetype = "dashed",
             color = "#E05A1E", linewidth = 1) +
  geom_vline(xintercept = 30, linetype = "dashed",
             color = "#C0392B", linewidth = 1) +
  annotate("text", x = 20.5, y = Inf, label = "High\nthreshold",
           vjust = 1.5, hjust = 0, size = 3.5, color = "#E05A1E") +
  annotate("text", x = 30.5, y = Inf, label = "Very High\nthreshold",
           vjust = 1.5, hjust = 0, size = 3.5, color = "#C0392B") +
  scale_x_continuous(limits = c(0, 100)) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Distribution of Max Medication Orders in Any 60-Minute Window",
    subtitle = "Per ICU admission | MIMIC-IV",
    x = "Max Orders in 60-Minute Window", y = "Number of Admissions",
    caption = "X-axis truncated at 100; max observed = 432. Median = 18."
  ) +
  theme_minimal(base_size = 13)
```

## Burst Intensity by Category

```{r plot-burst-category, fig.height=4, fig.cap="Figure 11. Percentage of ICU admissions by 60-minute burst intensity category. Thresholds: Low <10, Moderate 10-19, High 20-29, Very High 30+ orders per hour. MIMIC-IV."}
burst_cat_summary <- burst_metric %>%
  group_by(burst_category) %>%
  summarise(n = n(), pct = round(n() / nrow(burst_metric) * 100, 1),
            .groups = "drop")

ggplot(burst_cat_summary, aes(x = burst_category, y = pct, fill = burst_category)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = paste0(pct, "%")),
            vjust = -0.5, size = 4.5, fontface = "bold") +
  scale_fill_manual(values = c("#70AD47","#2E75B6","#E05A1E","#C0392B")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "ICU Admissions by 60-Minute Burst Intensity Category",
    subtitle = "MIMIC-IV ICU cohort",
    x = NULL, y = "Percent of Admissions"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

------------------------------------------------------------------------

# Preprocessing Notes

The following decisions were made during data preparation. Each is
documented with the reason it was made so the analysis can be understood
and reproduced.

**Lab events queried via DuckDB** — The labevents file in MIMIC-IV
contains tens of millions of rows, exceeding available RAM. DuckDB was
used to run SQL queries directly on disk, pulling only rows belonging to
patients in the ICU cohort without loading the full file into memory.

**Prescriptions loaded with `select = c(...)`** — Only the five columns
needed for analysis were loaded from the prescriptions file, reducing
RAM usage by approximately 80% compared to loading all 26 columns.

**Prescription counts aggregated before joining** — Medication orders
were summarized at the admission level before merging with the ICU
cohort. Joining raw prescription records directly would have multiplied
the number of rows in the dataset by hundreds, making it unmanageable.

**ICU units with n \< 200 collapsed to "Other"** — Eight ICU unit types
had fewer than 200 stays. Units this small cause convergence failure in
negative binomial estimation because the model cannot reliably estimate
coefficients from so few observations.

**Lab counts winsorized at the 99th percentile (3,086)** — The maximum
lab event count was 19,824. Values this extreme caused the negative
binomial model to fail. Winsorizing replaced the top 1% of values with
the 99th percentile cap, leaving 99% of the data unchanged.

**Rate variables derived (events per ICU day)** — Raw counts are
confounded by length of stay. A patient staying 30 days accumulates more
events than a 2-day patient regardless of care intensity. Dividing by
ICU days produces a fair per-day rate for comparison.

**Negative binomial over Poisson** — Confirmed by likelihood ratio test:
Chi-sq = 8,224,010, p ≈ 0. Variance-to-mean ratios of 119 to 1,178
across outcomes confirm severe overdispersion (see Table 4 and Table 5).

**Cohort lookup deduplicated before temporal join** — Some hospital
admissions contain multiple ICU stays, giving multiple intime values per
hadm_id. A deduplicated lookup table was created before joining
prescription timestamps to prevent row multiplication in the temporal
analysis.

**Burst thresholds recalibrated to 10/20/30** — The original thresholds
(2/5/10) classified 80.7% of admissions as High because the median burst
in this dataset is 18 orders per hour. Recalibrated thresholds produce a
meaningful spread across all four categories (see Figure 11).

```{r Descriptive-Summary}

# Medication orders summary
cohort %>%
  summarise(
    mean_rx     = round(mean(n_prescriptions, na.rm = TRUE), 1),
    median_rx   = round(median(n_prescriptions, na.rm = TRUE), 1),
    sd_rx       = round(sd(n_prescriptions, na.rm = TRUE), 1),
    mean_rxpd   = round(mean(rx_per_day, na.rm = TRUE), 1),
    median_rxpd = round(median(rx_per_day, na.rm = TRUE), 1)
  )

# Lab events summary
cohort %>%
  summarise(
    mean_lab     = round(mean(n_lab_events, na.rm = TRUE), 1),
    median_lab   = round(median(n_lab_events, na.rm = TRUE), 1),
    sd_lab       = round(sd(n_lab_events, na.rm = TRUE), 1),
    mean_labpd   = round(mean(labs_per_day, na.rm = TRUE), 1),
    median_labpd = round(median(labs_per_day, na.rm = TRUE), 1)
  )

# Total events summary
cohort %>%
  summarise(
    mean_total     = round(mean(total_events, na.rm = TRUE), 1),
    median_total   = round(median(total_events, na.rm = TRUE), 1),
    mean_totalpd   = round(mean(total_events_per_day, na.rm = TRUE), 1),
    median_totalpd = round(median(total_events_per_day, na.rm = TRUE), 1)
  )
```
